---
title: "Linear models are (almost) all you need to know"
subtitle: "Advanced Statistics and Data Analysis"
author: "Davide Risso"
format: 
    beamer:
        aspectratio: 169
        include-in-header: template.tex
        toc: true

knitr:
    opts_chunk:
        out-width: 75%
        fig-align: center
---

# ANOVA

## Analysis of Variance (ANOVA)

Last semester you have seen ANOVA as a generalization of the t-test for more than two groups.

Let's recall it with an example.

```{r, echo=TRUE}
head(mtcars)
```

## ANOVA example

Does car efficiency, measured in miles-per-gallon (mpg), depend on the type of engine?

In our data, we have three engyine types: 4, 6, and 8 cylinder engines.

We can use a one-way ANOVA to test whether the efficiency is the same or different across cars with these engines.

## ANOVA example

```{r}
library(tidyverse)

```

# The t-test and other tests

Consider introducing ranks and show that the nonparametric tests can also be written as linear models (perhaps as exercise/homework).

# Contrasts

# Interactions

