---
title: "Generalized Linear Models"
subtitle: "Advanced Statistics and Data Analysis"
author: "Davide Risso"
format: 
    revealjs:
        theme: default
        incremental: true
        scrollable: true
html-math-method:
    method: mathjax
    url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
knitr:
    opts_chunk:
        out-width: 75%
        fig-align: center
---

# Introduction

## Linear models

Linear models are the most useful applied statistics technique.

However, they have their limitations:

- The response may not be continuous
- The response may have a limited range of possible values

. . .

In such cases, linear models are not ideal and we may want to consider _generalizations_ of it.

## Generalized Linear Models (GLMs)

The **Generalized Linear Model (GLM)** is a family of models that _includes_ the linear model and extend it to situations in which the response variable has a distribution of the _exponential family_.

. . .

The _exponential family_ is a family of distributions that includes the Gaussian, Poisson, Binomial, Gamma, among others.

## Exponential Family

The density function of a distribution of the exponential family can be written as
$$
f(y_i; \theta_i, \phi) = \exp \left\{\frac{y_i \theta_i - b(\theta_i)}{a(\phi)} + c(y_i, \phi)\right\},
$$
where:

- $\theta_i$ are the _canonical parameters_;
- $\phi$ is the _dispersion parameter_;
- $a$, $b$, and $c$ are functions that depend on the specific distribution.


## Example: Poisson distribution {.smaller}

The probability function of a Poisson distribution can be written as
$$
f(y_i; \lambda_i) = \frac{\lambda_i^{y_i} e^{-\lambda_i}}{y_i!},
$$

or equivalently as
$$
f(y_i; \lambda_i) = \exp\left\{y_i \log(\lambda_i) - \lambda_i - \log(y_i!)\right\},
$$

which corresponds to an exponential family with:

- $\theta_i = \log(\lambda_i)$;
- $b(\theta_i) = \exp(\theta_i)$;
- $a(\phi) = 1$;
- $c(y_i, \phi) = -\log(y_i!)$.

## Log-likelihood function {.smaller}

Note that the log-likelihood function of the exponential family can be written as
$$
\ell(\lambda; y) = \sum_{i=1}^n \frac{y_i \theta_i - b(\theta_i)}{a(\phi)} + c(y_i, \phi).
$$

One can show that:
$$
E[Y_i] = \mu_i = b'(\theta_i)
$$
and
$$
Var(Y_i) = b''(\theta_i) a(\phi),
$$

where $b'()$ and $b''()$ are the first and second derivative of $b$.

::: {.callout-important}
Note that the variance depends on the mean!
:::

## Example: Poisson distribution

In the case of the Poisson, $b(\theta_i) = \exp(\theta_i)$ and $a(\phi) = 1$, hence:

$$
E[Y_i] = \exp(\theta_i) = \lambda_i
$$
and

$$
Var(Y_i) = \exp(\theta_i) = \lambda_i.
$$

## Generalizations of the linear model

We can think of GLMs as a two-way generalization of linear models:

1. We can use a _function_ of the linear predictor to transform it to a range of values that is more useful for the application.

2. We can use a different _distribution_ for the error terms other than the Gaussian.

## Likelihood estimation

Recall that ordinary least squares can be seen as the _maximum likelihood estimate_ of the linear regression parameters.

. . .

The idea is that by changing the distribution of the response variable, we can still use maximum likelihood to estimate the regression coefficients, but specifying a _different likelihood function_.

## The three components of a GLM

The three components that we need in order to fully specify a GLM are:

1. An _exponential family_ distribution for the response.

2. A _link function_ that connects the linear predictor to the response.

3. A _variance function_ that specifies how the variance depends on the mean.

## Example: the normal linear model

If we assume that:

1. The response variable is distributed as a Gaussian r.v.

2. The link function is the _identity function_.

3. The variance is _constant_ with respect to the mean.

. . . 

Then, we have the standard linear model.

## Example: the normal linear model

In formulas:

1. $Y \sim \mathcal{N}( \mu, \sigma^2)$.

2. $E[Y | X] = X \beta$.

3. $V(\mu) = \sigma^2$.

## Example: logistic regression

Logistic regression assumes that:

1. The response variable is distributed as a Bernoulli r.v.

2. The link function is the _logistic function_.

3. The variance is _a function_ of the mean.

## Example: logistic regression

In formulas:

1. $Y \sim \mathcal{Be}(\mu)$.

2. $g(E[Y | X]) = X \beta$, where $g(\mu) = \log \left(\frac{\mu}{1-\mu}\right)$.

3. $V(\mu) = \mu (1 - \mu)$.

. . .

Note that the link function implies that
$$
\mu = \frac{\exp(X \beta)}{1 + \exp(X\beta)}.
$$

## Example: logistic regression

- Assume that we have collected data from 299 patients that had heart failure.

- For each patient we measure if they died or survived after a follow-up period.

- We can use a set of 13 clinical features to try to predict patient survival.

```{r}
library(tidyverse)
theme_set(theme_minimal(base_size = 20))

heart <- read_csv("data/heart_failure_clinical_records_dataset.csv")
```

- In total, `r sum(heart$DEATH_EVENT)` patients out of `r nrow(heart)` died.


## Example: logistic regression

```{r}
ggplot(heart, aes(x = time, y = DEATH_EVENT)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE)
```

## Example: logistic regression

```{r}
ggplot(heart, aes(x = time, y = DEATH_EVENT)) +
    geom_point() +
    geom_smooth(method = "glm", method.args = list(family=binomial), se = FALSE)
```


## Example: Poisson regression

Poisson regression (or log-linear regression) assumes that:

1. The response variable is distributed as a Poisson r.v.

2. The link function is the _log function_.

3. The variance is _a function_ of the mean.

## Example: Poisson regression

In formulas:

1. $Y \sim \mathcal{Poi}(\mu)$.

2. $g(E[Y | X]) = X \beta$, where $g(\mu) = \log(\mu)$.

3. $V(\mu) = \mu$.

. . .

Note that the link function implies that
$$
\mu = \exp(X \beta)
$$

::: {.callout-important appearance="minimal"}
The log is called the _canonical link function_ because it is equal to the function that transforms the mean into the _canonical parameter_ in the exponential family formulation.
:::

## Example: Poisson regression

- We measure the expression of two genes, DUSP1 and VCAM1, in airway smooth muscle (ASM) cell lines.

- The DUSP1 gene plays an important role in the human cellular response to environmental stress.

- The VCAM1 gene is involved in cell adhesion and signal transduction.

- The genes are measured with an assay that yields counts as a measure of gene expression.

## Example: Poisson regression

```{r}
library(airway)
data("airway")
y <- assay(airway)["ENSG00000120129",]
df <- as.data.frame(colData(airway))
df$DUSP1 <- y
y <- assay(airway)["ENSG00000162692",]
df$VCAM1 <- y


ggplot(df, aes(x = VCAM1, y = DUSP1)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE)
```

## Example: Poisson regression

```{r}
library(airway)
data("airway")
y <- assay(airway)["ENSG00000120129",]
df <- as.data.frame(colData(airway))
df$DUSP1 <- y
y <- assay(airway)["ENSG00000162692",]
df$VCAM1 <- y


ggplot(df, aes(x = VCAM1, y = DUSP1)) +
    geom_point() +
    geom_smooth(method = "glm", method.args = list(family=poisson), se = FALSE)
```

## Quasi Likelihood {visibility="hidden"}

To completely define a GLM it is sufficient to specify:

- The link function:
    - i.e., how the linear predictor relates to the conditional mean;
- The variance function:
    - i.e., the relationship between the mean and the variance.
    
. . .

Specifying these two components gives us a way to fit the GLM even if there is no proper distribution of $Y$. In such cases, we talk about _quasi-likelihood_.

# Parameter estimation

## Log-likelihood of a GLM

In principle, we can maximize the log-likelihood as a function of the regression coefficients.

In fact, once we have defined a relationship between the parameter of the model and the predictors, we can write the likelihood as a function of $\beta$.

. . .

In the case of the Poisson, the log-likelihood is
$$
\ell(\beta; X, y) = \sum_{i=1}^n y_i \eta_i- \exp\{\eta_i\},
$$
where $\eta_i = \log(\mu_i) = \sum_{k=1}^p x_{ik} \beta_k$.

## Score function {.smaller}

The derivative of the log-likelihood with respect to $\theta$ is also called _the score function_.

One can show that in the exponential family, the score function
$$
S(\theta) = \sum_{i=1}^n \frac{y_i - \mu_i}{a(\phi)},
$$

. . .

when you want to derive for $\beta$, the chain rule of differentiation gives
$$
S(\beta) = \sum_{i=1}^n \frac{Y_i - \mu_i}{Var(Y_i)} \frac{d \mu_i}{d \eta_i} X^\top = X^\top A (Y - \mu),
$$

where $A$ is a diagonal matrix with elements $a_{ii} = (Var(Y_i) \frac{d \eta_i}{d \mu_i})^{-1}$.

::: {.callout-important}
Note the relation with least squares!
:::

## Poisson example

In Poisson data:

- $Var(Y_i) = E[Y_i] = \mu_i$;
- $\eta_i = \log(\mu_i)$, hence $\frac{d \eta_i}{d \mu_i} = \frac{1}{\mu_i}$.

. . .

Hence, $A = I$, the identity matrix and
$$
S(\beta) = X^\top (Y - \mu).
$$

## Derivation {.smaller} {visibility="hidden"}

The log-likelihood is
$$
\ell(\lambda; y) = \sum_{i=1}^n \frac{y_i \theta_i - b(\theta_i)}{a(\phi)} + c(y_i, \phi).
$$

And the score is
\begin{align*}
S(\beta) = \frac{\partial \ell}{\partial \beta} &= \sum_{i=1}^n \frac{d \ell_i}{d \theta_i} \frac{d \theta_i}{d \mu_i} \frac{\partial \mu_i}{\partial \beta} \\
&= \sum_{i=1}^n \frac{y_i - b'(\theta_i)}{a(\phi)} \frac{1}{b''(\theta)} \frac{\partial \mu_i}{\partial \beta}.
\end{align*}

Since $b''(\theta_i) = \frac{d \mu_i}{d \theta_i} = Var(Y_i)$,
$$
S(\beta) = \sum_{i=1}^n \frac{Y_i - \mu_i}{Var(Y_i)} \frac{d \mu_i}{d \eta_i} X^\top = X^\top A (Y - \mu)
$$

For canonical link functions, $\theta_i = x_i \beta$, hence $A=I$ and
$$
S(\beta) = X^\top (Y - \mu)
$$

## Solving the score equations

To find the MLE we need to solve the equation
$$
S(\beta) = X^\top A (Y - \mu) = 0.
$$

The problem is that this is non linear in $\beta$ because of the link function.

We hence need numerical algorithms to find the values of $\beta$ that solve this equation.

## Newton-Raphson

```{r}
S <- function(beta) (beta - 3)^2 
dS <- function(beta) 2*(beta-3)
dSline <- function(beta, betak) dS(betak) *beta + S(betak) - dS(betak) * betak

beta <- seq(-21,15,length=100)
betak <- -20
betak1 <- betak - 1/dS(betak)*S(betak)
betaSeqDk <- seq(betak,betak1,length=100)
pNR <- qplot(beta, S(beta), geom="line") +
  annotate("line",x=betaSeqDk,y=dSline(betaSeqDk,betak),linetype = "dashed",col="red") +
    annotate("line",x=rep(betak1,2),y=c(0,S(betak1)),linetype = "dashed",col="red") +
    annotate("point",x=betak,y=S(betak),col="red") +
  theme_bw()
betaNs <- betak1
for (i in 1:5)
{
betak <- betak1
betak1 <- betak - 1/dS(betak)*S(betak)
betaNs <- c(betaNs,betak1)
betaSeqDk <- seq(betak,betak1,length=100)
pNR <- pNR +
    annotate("line",x=betaSeqDk,y=dSline(betaSeqDk,betak),linetype = "dashed",col="red") +
    annotate("line",x=rep(betak1,2),y=c(0,S(betak1)),linetype = "dashed",col="red") 
}
pNR + 
  annotate("text",x=-20,y=S(-20)+40,label=paste0("S(beta[k])"),col="red",parse=TRUE) +
  annotate("text",x=betaNs[1:2],y=S(betaNs[1:2])+40,label=paste0("S(beta[k+",1:2,"])"),col="red",parse=TRUE)   +
  annotate("text",x=betaNs[1:3],y=rep(-40,3),label=paste0("beta[k+",1:3,"]"),col="red",parse=TRUE) +
  annotate("point",x=betaNs,y=S(betaNs),col="red") +
  annotate("text",x=-20,y=400, label="frac(partialdiff*S, partialdiff*beta)*' |'*beta[k]", col="red",parse=TRUE) +
  xlab(expression(beta)) +
  ylab(expression(S(beta)))

```

## Newton-Raphson

The Newton-Raphson algorithm is a numerical algorithm to find the "root" of a function, i.e., to find the value for which that function is equal to 0.

. . . 

It starts from an initial guess, say $\beta_k$, and moves in the direction of the derivative until it intersects the zero line.

## Newton-Raphson

Specifically:

1. Choose an initial parameter estimate $\beta_k$;
2. Calculate the score $S(\beta_k)$;
3. Calculate the derivative $\frac{\partial S(\beta)}{\partial \beta}$ at the value $\beta_k$;
4. Go in the direction of the derivative until the point, $\beta_{k+1}$ in which it crosses the zero line;
5. Update the value of $\beta$ to $\beta_{k+1}$;
6. Iterate 2-5 until convergence.

## Derivative of the score

For the Newton-Raphson algorithm to work, we need to compute the derivative of the score function.

One can show that it is 
$$
\frac{\partial S(\beta)}{\partial \beta} = - X^\top W X,
$$
where $W$ is a diagonal matrix with elements
$$
w_{ii} = \frac{(d \mu_i / d \eta_i)^2}{Var(Y_i)}
$$

## Iteratively Reweighted Least Squares {.smaller}

Now that we know the derivative, we can rewrite the Newton-Raphson algorithm as the _Iteratively Reweighted Least Squares_ (IRLS).

In fact we can show that
$$
\beta_{k+1} = \beta_k + (X^\top W X)^{-1} S(\beta_k)
$$
or equivalently
$$
\beta_{k+1} = (X^\top W X)^{-1} X^\top W Z,
$$
where
$$
Z = X \beta_k + \frac{\partial \eta}{\partial \mu} (Y - \mu).
$$

## Poisson example {visibility="hidden"}

For Poisson data, we can see that

- $\frac{\partial \eta}{\partial \mu} = \frac{\partial \log(\mu)}{\partial \mu} = \frac{1}{\mu}$
- $W$ is a diagonal matrix with $w_{ii} = \mu_i$

## Interpretation of IRLS

Note that the estimate of $\beta$ at iteration $k$ is similar to the least squares estimator, but it is weighted by a function of the variance.

This is a way to account for the heteroschedasticity of the data.

## Example: Poisson regression {visibility="hidden"}

Let's go back to the example of the two genes.

We can fit the model:
$$
\log E[Y | X] = \beta_0 + \beta_1 X,
$$
where $Y$ is the expression of DUSP1 and $X$ is the expression of VCAM1.

## Example: Poisson regression {.smaller} {visibility="hidden"}

We can now iteratively fit the model by:

1. Choose initial values of $\beta^0 = [\beta_0 \, \beta_1] = [0 \, 1]$
2. Calculate the score $S(\beta^0) = $

## In R: the `glm()` function

In R there is a very convenient function that implements the IRLS algorithm and has a syntax very similar to the `lm()` function.

In addition to the formula, you need to specify the `family` of distributions you want to specify for the response variable and (implicitly or explicitly) the link function.

```{r}
#| echo: true
#| eval: false

fit <- glm(y ~ x, family = poisson)

fit <- glm(y ~ x, family = poisson(link = "log"))
```


## Example: Poisson regression

In the airway smooth muscle (ASM) cell lines, we have two additional variables to consider:

1. The cell lines were either untreated or treated with dexamethasone (`dex`).

2. The cell lines were derived from four distinct donors (`cell`).

. . .

It is of interest to check whether the DUSP1 gene changes between treated and control cells, potentially accounting for the differences in cell lines.

## Example: Poisson regression

```{r}
df$dex <- relevel(df$dex, ref="untrt")
fit <- glm(DUSP1 ~ dex, family = poisson, data = df)
fit
```

## Interpretation of the parameters {.smaller}

The Poisson regression model can be written as
$$
\log \mu_i = \beta_0 + \beta_1 x_{i},
$$
where $x_i=1$ if the cells are treated and $x_i=0$ if untreated.

Hence,
$$
\log(\mu_i | x_i = 1) - \log(\mu_i | x_i = 0) = \beta_1
$$

which leads to
$$
\beta_1 = \log \left(\frac{\mu_i | x_i = 1}{\mu_i | x_i = 0}\right).
$$

## Example: logistic regression

In the heart failure dataset, we may be interested in the relation between diabetes and heart failure.

## Example: logistic regression

```{r}
fit <- glm(DEATH_EVENT ~ diabetes, family = binomial, data = heart)
fit
```

## Interpretation of the parameters {.smaller}

The logistic regression model can be written as
$$
\log \left( \frac{\mu_i}{1-\mu_i} \right)= \beta_0 + \beta_1 x_{i},
$$
where $x_i=1$ if the subject has diabetes and $x_i=0$ if not.

Hence,
$$
\log \left( \frac{\mu_1}{1-\mu_1} \right) - \log \left( \frac{\mu_0}{1-\mu_0} \right) = \beta_1,
$$
where we indicate with $\mu_1 = E[Y|x=1]$ and $\mu_0 =  E[Y|x=0]$.

We can thus interpret $\beta_1$ as the _log odds ratio_
$$
\beta_1 = \log OR = \log \left(\frac{\mu_1 / (1-\mu_1)}{\mu_0 / (1-\mu_0)}\right).
$$

# Inference

## Likelihood Ratio Test

The analogous of the sum of squares and the F test in GLMs is the log-likelihood and the likelihood ratio test.

. . .

We can use the likelihood to compare two models, as we have done with the F statistics in linear models.

. . .

One can show that, when $n \to \infty$,
$$
D = 2 [\ell(\hat\beta) - \ell(\hat\beta^0)] \stackrel{H_0}{\sim} \chi^2_{df-df_0},
$$

where $\hat\beta$ and $\hat\beta^0$ are the estimates of the coefficient under the full and reduced model, with $df$ and $df_0$ degrees of freedom, respectively.

## Wald Test {.smaller}

An alternative to the LRT is the Wald test, that leverages the fact that when $n \to \infty$, 
$$
\hat\beta \stackrel{H_0}{\sim} \mathcal{N}(\beta, (X^\top W X)^{-1}).
$$

Hence, the statistic
$$
t = \frac{\hat\beta_k}{\sqrt{[(X^\top W X)^{-1}]_{kk}}} \stackrel{H_0}{\sim} N(0, 1)
$$
can be used to compute confidence intervals and test the null hypothesis $\beta_k = 0$.

## Example: Poisson regression

In the airway smooth muscle (ASM) cell lines, we have two additional variables to consider:

1. The cell lines were either untreated or treated with dexamethasone (`dex`).

2. The cell lines were derived from four distinct donors (`cell`).

. . .

It is of interest to check whether the DUSP1 gene changes between treated and control cells, potentially accounting for the differences in cell lines.

## Poisson regression: Wald test

```{r}
fit <- glm(DUSP1 ~ dex, family = poisson, data = df)
summary(fit)
```

## Poisson regression: Wald Test

```{r}
fit2 <- glm(DUSP1 ~ dex + cell, family = poisson, data = df)
summary(fit2)
```

## Poisson regression: LRT

```{r}
anova(fit, fit2)
```

## Logistic regression: 

```{r}
fit <- glm(DEATH_EVENT ~ . - time, family = binomial, data = heart)
summary(fit)
```

## Logistic regression: Wald test

```{r}
fit0 <- glm(DEATH_EVENT ~ age + sex + smoking, family = binomial, data = heart)
anova(fit0, fit)
```
