---
title: "Bayesian inference"
subtitle: "Advanced Statistics and Data Analysis"
author: "Davide Risso"
format: 
    revealjs:
        theme: default
        incremental: true
        scrollable: true
html-math-method:
    method: mathjax
    url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
knitr:
    opts_chunk:
        out-width: 75%
        fig-align: center
---

# Introduction

## The Frequentist Phylosophy

The statistical methods discussed so far are known as _frequentist_.

The frequentist point of view is based on the following ideas:

1. **Probabilities are objective properties of the real world**: they are defined as the limit of the relative frequencies in infinite many trials.

2. **Parameters are fixed, unknown constants**.

. . .

As a consequence, a frequentist methods are designed to have good _long-run frequency properties_.

## Example

- A $95\%$ confidence interval is defined so that _it includes the true value of the parameter $95\%$ of the times if we repeat the experiment an infinite number of times_.

- **Note that it is not the parameter but the interval to be random!**

## The Bayesian Phylosophy

The Bayesian approach takes a very different point of view:

1. **Probabilities describe a degree of belief**: they are not objective properties of the world, but rather _subjective_ propositions.

2. **Parameters are random variables**: there is no such thing as the true parameter, all we know is the probability distribution of its values.

. . .

As a consequence, inference in a Bayesian setting takes the form of an "updated" distribution of the parameter: we can then extract point estimates and intervals from this distribution.

## Model specification

- In frequentist inference, we need a model for the data generating distribution (i.e., the likelihood).

- In the Bayesian framework, in addition to the likelihood, we need to specify a **prior distribution** for the parameter(s).


- This prior distribution _elicit_ our prior belief in what the value of the parameter should be.

- Inference is carried out by updating the prior distribution with the new knowledge that we observed in the data, to form the **posterior distribution**.

# Bayes' Theorem

## Probability

A fair coin is tossed twice. What is the probability of observing
exactly one Head?

. . .

The sample space is $S=\{HH,HT,TH,TT\}.$ Because the coin is fair, each
of the four outcomes in $S$ is equally likely. 

. . .

Let $A$ denote the event
that exactly one Head is observed. Then, $A = \{HT, TH\},$ and
$$P(A) = \frac{\#(A)}{\#(S)} = \frac{2}{4} = 0.5.$$


## Conditional Probability {.smaller}

A fair coin is tossed twice. Suppose that we toss the coin once and we observe Tail. What is _now_ the probability of observing exactly one Head?

. . .

We denote with $B$ the event that we observed (that the first toss is Tail), and with $A$ the event that exactly one Head is observed. 

. . .

As before, $A = \{HT, TH\}$, but we have $B = \{TH, TT\},$ so the only event in $A$ that is compatible with $B$ is $\{TH\}$.

. . .

We hence computed the _conditional probability_ of the event 
$$
P(A | B) = \frac{\#(A \cap B)}{\#(S \cap B)} = \frac{1}{2} = 0.5.
$$

. . .

What would be the conditional probability if the first coin was Head?

## Conditional Probability

<span>*If $A$ and $B$ are events, and $P(B) > 0$, the conditional
probability of $A$ given $B$ is*</span>
$$P(A|B) = \frac{P(A\cap B)}{P(B)} .$$ 

. . .

The same equality can be written
as: $$P(A|B) P(B)= P(A\cap B).$$

## Example: COVID-19 test {.incremental}

**Thanks to Laura Ventura**

Suppose that we perform a test to check whether we have contracted COVID-19. The possible
observable outcomes are two: we have contracted the
virus $(D),$ or we have not $(H)$. 

. . .

From the latest statistical analyses, we can estimate the *prevalence*
of the disease in the population, $P(D)$, i.e.,
the probability that an individual selected at random from the population is infected.

## Diagnostic Test {.smaller}

A test is designed to detect the presence of the SARS-CoV-2 virus in the nose and throat. This test also has two possible outcomes: positive for the presence of viral RNA $(+),$ negative for the presence of viral RNA $(-)$.

. . . 

Because diagnostic procedures undergo extensive evaluation before they
are approved for general use, we
have a fairly precise notion of the probabilities of a *false
positive*, i.e., the probability of obtaining a positive test
result given that the patient is not infected, and a *false
negative*, i.e., the probability of obtaining a negative test
results given that the patient is infected.

. . .

These probabilities are conditional probabilities: the probability of a
false positive is $P(+|H),$ and the probability of a false negative is
$P(-|D)$.

## Predictive Value of the Test {.smaller}

The gold standard test for COVID-19 has a false positive and false negative rate of about $5\%$.

. . .

Assume that there is a COVID-19 prevalence of about $8\%$ in the Italian population.

. . .

Assume $P(D) = .08,$ $P(+|H) = .05$ and $P(-|D) = .05$.

. . .

(This implies $P(H) = .92,$ $P(-|H) = .95$ and $P(+|D) = .95$.)

. . .

What is the <span>*predictive*</span> value of the test, i.e., the
probability that we don't have COVID-19 if the test is negative?

. . .

The question asks to compute the quantity $P(H|-).$ By definition, we
have $$P(H|-) = \frac{P(H\cap -)}{P(-)} = \frac{P(-|H) P(H)}{P(-)}.$$

## Computations {.smaller}

We know that $$P(H \cap -) = P(H) P(-|H) = 0.92 \times 0.95,$$

. . .

$$P(D \cap -) = P(D) P(-|D) = 0.08 \times 0.05,$$

. . .

$$P(H \cap +) = P(H) P(+|H) = 0.92 \times 0.05,$$

. . .

$$P(H \cap -) = P(H) P(-|H) = 0.92 \times 0.95,$$

. . .

$$P(-) = P(D \cap -) + P(H \cap -)=0.08 \times 0.05 + 0.92 \times 0.95.$$

. . .

Therefore,
$$P(H|-) = \frac{P(H\cap -)}{P(-)} = \frac{0.92 \times 0.95}{0.08 \times 0.05 + 0.92 \times 0.95}=0.995.$$

## Bayes' Theorem

Note that we have just used one of the most important theorems in modern statistics, _Bayes' Theorem_.

. . .

The theorem states:
$$
P(B | A) = \frac{P(A | B) P(B)}{P(A)}.
$$

. . .

In words, if we know the conditional probability of $A$ given $B$, in addition to the marginal probabilities of the two events, we have a rule to compute the conditional probability of $B$ given $A$.

## Example: births in Paris

- Laplace was the first to formulate a statistical problem in Bayesian terms.

- The question he poses was whether the probability of a female birth is or is not lower than 0.5.

- Data: in Paris, from 1745 to 1770 there were 493,472 births, of which 241,945 were girls.

## A frequentist solution

- We can assume a binomial model and get a frequentist estimate for the probability of a female birth as $\hat\theta = \frac{241,945}{493,472} = 0.4903$

- We can also compute a confidence interval $[0.4889, 0.4917]$

- Finally, we can test the null hypothesis $H_0: \theta = 0.5$ and reject it as the obtained p-value is close to 0.

## A Bayesian solution

- Laplace argues that he has no reason to _a priori_ believe that any value of $\theta$ is more likely than others.

- Hence, the _prior probability_ is $p(\theta) \sim \mathcal{U}(0, 1)$.

- We can prove that in such case the _posterior distribution_ is 

. . .

$$
p(\theta | \text{data}) \propto \theta^{241,945} \, (1-\theta)^{493,472 - 241,945}
$$

## The posterior distribution

```{r}
library(tidyverse)
theme_set(theme_minimal(base_size = 20))
df <- data.frame(theta = c(0, 1))
ggplot(df, aes(theta)) +
    stat_function(fun = dbeta, n = 101, args = list(shape1 = 1, shape2 = 1), aes(color="Prior")) +
    stat_function(fun = dbeta, n = 101, args = list(shape1 = 24+1, shape2 = 49-24+1), aes(color="Posterior"))
```


## Bayes' Theorem revisited {.smaller}

In statistics, we can rethink the Bayes' Theorem in terms of the data and the parameter:
$$
p(\theta | y) = \frac{p(\theta) \, p(y | \theta)}{p(y)},
$$
where:

- $p(\theta)$ is the _prior distribution_ of the parameter
- $p(y | \theta)$ is the distribution of the data given a particular value of the parameter (i.e., _the likelihood_)
- $p(y)$ is the _marginal distribution of the data_ (across all possible values of the parameter)

. . .

::: {.callout-note}
Much of Bayesian statistics deals with ways to compute/approximate $p(y)$.
:::


## Example: Haplotype frequencies

- A _haplotype_ is a collection of alleles (DNA sequence variants) that are spatially adjacent on a chromosome, are usually inherited together (recombination tends not to disconnect them), and thus are genetically linked. In this case we are looking at linked variants on the Y chromosome.

- We are interested in the frequencies of particular Y-haplotypes that consist of a set of different short tandem repeats (STR). The combination of STR numbers at the specific locations used for DNA forensics are labeled by the number of repeats at the specific positions. 

## Example: Haplotype frequencies

The data look like this:

```{r}
haplo6 <- read.table("data/haplotype6.txt", header = TRUE)
haplo6
```

For instance, haplotype H1 has 14 repeats at position DYS19, 12 repeats at position DXYS156Y, etc.

Suppose we want to find the underlying proportion $\theta$ of a particular haplotype in a population of interest, by haplotyping $n=300$ men; and suppose we found H1 in $y = 40$ of them.

## Frequentist approach

We can assume that the data generating distribution is $Y \sim \mathcal{Bin}(n, \theta)$, with $\theta$ fixed and unknown.

. . .

By now we know that the _maximum likelihood estimate_ of $\theta$ is $\hat{\theta} = 40/300 = 0.13$.

. . .

We can compute confidence intervals and test hypotheses on the true value $\theta$.

## Bayesian approach

- In a Bayesian setting, $\theta$ is not fixed, and we need to assume a _prior_ probability distribution for the parameter as well.

- In principle we could use _any_ distribution.

- It is convenient to specify a _Beta distribution_, i.e.

. . .

$$
p_{\alpha, \beta}(x) = \frac{x^{\alpha-1} (1-x)^{\beta-1}}{B(\alpha, \beta)},
$$
where $B(\alpha, \beta) = \frac{\Gamma(\alpha) \Gamma(\beta)}{\Gamma(\alpha+\beta)}$ is called the _beta function_.


## Prior distribution

- We can see how the prior distribution depends on two parameters, in this case $\alpha$ and $\beta$.

- These are known as _hyper-parameters_ and are typically chosen according to some prior information.

- The choice of a Beta distribution in this case is mathematically convenient, as one can show that the posterior distribution will be again a Beta distribution.

- We say that the Beta distribution is the _conjugate prior_ for the binomial model.

## How to choose the hyperparameters

- The hyperparameters are our way to express our prior knowledge of the problem.

- An intuitive way is to think about the prior mean and variance as a way to elicit what is the value that we expect before performing the data (mean) and how confident we are in our prediction (variance).

- The Beta distribution is quite flexible in that it includes the _uniform distribution_ (absence of prior information) as well as very concentrated distributions.

## How to choose the hyperparameters

If $X \sim \mathcal{Beta}(\alpha, \beta)$, then
$$
E[X] = \frac{\alpha}{\alpha + \beta} 
$$
and

$$
Var(X) = \frac{\alpha\beta}{(\alpha + \beta)^2(\alpha+\beta+1)}
$$

## Example: Haplotype frequencies

Let's think about three situations:

- We have no prior knowledge on the haplotype frequencies in the population: we can choose a uniform distribution.

- We know from the literature that one in five men should have haplotype H1, but we are not very confident.

- We are very confident that one in five men has haplotype H1.

## Example: Prior distribution

```{r}
ggplot(df, aes(theta)) +
    stat_function(fun = dbeta, n = 101, args = list(shape1 = 1, shape2 = 1), aes(color="Noninformative")) +
    stat_function(fun = dbeta, n = 101, args = list(shape1 = 4, shape2 = 16), aes(color="Weakly informative")) +
    stat_function(fun = dbeta, n = 101, args = list(shape1 = 16, shape2 = 64), aes(color="Strongly Informative"))
```

## The posterior distribution

If the prior is Beta and the likelihood is Binomial, the posterior distribution is still Beta, but with updated parameters:
$$
p(\theta | y) \sim \mathcal{Beta}(\alpha + y, \beta + n - y).
$$

. . .

Hence,
$$
E[\theta | y] = \frac{\alpha+y}{\alpha + \beta + n},
$$

which is a synthesis of the prior information and the information contained in the data.

## The posterior distribution

```{r}
ggplot(df, aes(theta)) +
    stat_function(fun = dbeta, n = 1001, args = list(shape1 = 16, shape2 = 64), aes(color="Prior")) +
    stat_function(fun = dbeta, n = 1001, args = list(shape1 = 16+40, shape2 = 64+300-40), aes(color="Posterior")) +
    geom_vline(xintercept = 40/300, linetype = 2)
```

## The posterior distribution

```{r}
ggplot(df, aes(theta)) +
    stat_function(fun = dbeta, n = 1001, args = list(shape1 = 16, shape2 = 64), aes(color="Prior")) +
    stat_function(fun = dbeta, n = 1001, args = list(shape1 = 16+40, shape2 = 64+300-40), aes(color="Posterior")) +
    geom_vline(xintercept = 40/300, linetype = 2)
```

- Note that we do not have a single answer for the "best" value of $\theta$ but a distribution.

- However, we can calculate the _posterior mean_ and _probabilities_ based on the posterior distribution, e.g., $Pr(\theta > 0.2) \approx 0.003$.


## The posterior distribution

Let's look again at the posterior distribution:
$$
\theta | y \sim \mathcal{Beta}(\alpha + y, \beta + n - y).
$$

- As we increase the sample size, we give more weight to the data.

- Conversely, if we increase the hyperparameters we give more weight to our prior belief.

## Credibility intervals

- The Bayesian analog of the _confidence intervals_ are the **credibility intervals**.

- These are much more natural to think about: they are straightforward to interpret in terms of the posterior probability of the parameter.

- For instance, we can use quantiles to compute the range in which the parameter lies with $95\%$ probability.

. . .

$$
Pr(q_{2.5} \leq \theta \leq q_{97.5}) = 0.95
$$

## Example: Haplotype frequencies

Putting it all together, we can specify our Bayesian model as:

- The data generating distribution is $y|\theta \sim \mathcal{Bin}(n=300, \theta);$

- The prior distribution is $\theta \sim \mathcal{Beta}(\alpha = 16, \beta = 64);$

. . .

Observing $y=40$ leads to the posterior distribution
$$
\theta | y \sim \mathcal{Beta}(16+40, 64 + 300 - 40).
$$

## Example: Haplotype frequencies

The posterior mean is

$$
E[\theta | y] = \frac{\alpha+y}{\alpha + \beta + n} = 0.147
$$

(compared to a MLE of $\hat\theta = 0.13$).

. . .

In addition, the 95% credibility interval is

```{r}
#| echo: true
qbeta(c(0.025, 0.975), shape1 = 16+40, shape2 = 64+300-40)
```

## Not just Beta

- Many of the concepts that we have seen in this example extend to many other distributions!

- For instance if we have Gaussian data, a Gaussian prior on the mean will lead to a Gaussian posterior.

- This is true for the linear model too!

## Not just conjugate priors

There are many different choices of prior distributions, and when closed form solutions do not exist to compute the posterior, we need to use numerical approximations or simulation approaches.

Markov Chain Monte Carlo (MCMC) and Variational Approximations are perhaps the two most popular techniques.













